The impact of cosmic variance on simulating weak lensing surveys

### 2014-11-11 Writeup/paper_v3.tex
minor tweaks

The intrinsic ellipticity distribution plays a role in nearly all
shear estimators.  In some, the role is explicit: for example, LensFit
\citep{2007MNRAS.382..315M,2008MNRAS.390..149K,2013MNRAS.429.2858M}
and the methods presented by \cite{2014MNRAS.438.1880B} require
accurate intrinsic ellipticity distributions as inputs, and
uncertainty in the distribution was one source of
systematic uncertainty in the CFHTLenS weak lensing results
\citep{2013MNRAS.432.2433H,2013MNRAS.429.2858M}.  In others, the role
is implicit.  For example, the re-Gaussianization method and several
other moment-based methods require calculation of a shear responsivity
\citep{BJ02,HS03} that describes how the galaxy population overall
responds to a shear, based on its intrinsic ellipticity distribution.
The responsivity can be calculated based on the observed shape
distribution, assuming that the uncertainties in the shears are known
enough that their contribution to that distribution can be removed.
If a simulated sample in some redshift slice has a different intrinsic
ellipticity distribution and therefore responsivity, it could lead to
incorrect conclusions about shear calibration.  The responsivity
scales roughly like $1-e_\text{RMS}^2$, which means that deviations in
RMS ellipticity at the level of 0.01 (Figs.~\ref{fig:rms_ellip} and~\ref{fig:rms_ellip_momentbased}) would become fractional shear
errors of
\begin{equation}
\frac{\Delta\gamma}{\gamma} \approx \frac{2 e_\text{RMS} \Delta
  e_\text{RMS} }{1-e_\text{RMS}^2} \approx 0.01.
\end{equation}
In the context of upcoming lensing surveys that seek to constrain
shears to the per cent level or better, a systematic error of this
magnitude in shear calibration is quite serious.

Regarding possible biases in the morphological mixtures of galaxies
due to overdensities or underdensities in the training sample, there
are results in the literature for several methods that show how shear
biases vary with morphology.  For example, for the maximum likelihood
fitting code {\sc im3shape}, figure 2 in \cite{2012MNRAS.427.2711K}
shows multiplicative biases for two-component \sersic profile galaxies
as a function of their bulge-to-total ratios (denoted there as
$F_b/(F_b+F_d)$, which we will equate with our $B/T$).  The shear
calibration bias scales roughly like $0.04 - 0.05(B/T)$ as $B/T$ goes
from $0$ to $1$.  Our results suggest that typical (median) $B/T$
values may be influenced by cosmic variance in the COSMOS field
leading to fluctuations of order $0.05$.  The resulting variation in
the calibration bias would therefore be $\sim 2.5\times 10^{-3}$, or
$0.25$ per cent shear calibration uncertainty.  For existing datasets
this is not very problematic, but for surveys like LSST, Euclid, and
WFIRST-AFTA, this would be a dominant part of the systematic error
budget.  To give another example, for re-Gaussianization, figure 9 of
\cite{2012MNRAS.420.1518M} shows that as \sersic $n$ goes from $1$ to
$6$, the shear calibration bias varies by $2$ per cent.  In this case,
since we have shown that the median value of \sersic $n$ can vary by
$\sim 0.4$ due to morphology-density correlations, this suggests that
the shear calibration for re-Gaussianization could be misestimated by
$\sim 2\times 10^{-3}$, or 0.2 per cent.  This too is acceptable in
existing datasets, but not those that will be used for shear
estimation in the next decade.

>>>>

The intrinsic ellipticity distribution plays a role in nearly all
shear estimators.  In some, the role is explicit: for example, LensFit
\citep{2007MNRAS.382..315M,2008MNRAS.390..149K,2013MNRAS.429.2858M}
and the methods presented by \cite{2014MNRAS.438.1880B} require
accurate intrinsic ellipticity distributions as inputs, and
uncertainty in the distribution was one source of
systematic uncertainty in the CFHTLenS weak lensing results
\citep{2013MNRAS.432.2433H,2013MNRAS.429.2858M}.  The intrinsic
ellipticity distribution enters the calculation for other methods in
other ways.  For example, the re-Gaussianization method and several
other moment-based methods require calculation of a shear responsivity
\citep{BJ02,HS03} that describes how the galaxy population overall
responds to a shear, based on its intrinsic ellipticity distribution.
The responsivity can be calculated based on the observed shape
distribution, assuming that the uncertainties in the shears are known
well enough that their contribution to that distribution can be removed.
If a simulated sample in some redshift slice has a different intrinsic
ellipticity distribution and therefore responsivity, it could lead to
incorrect conclusions about shear calibration.  The responsivity
scales roughly like $1-e_\text{RMS}^2$, which means that deviations in
RMS ellipticity at the level of 0.01 due to local environments (Figs.~\ref{fig:rms_ellip} and~\ref{fig:rms_ellip_momentbased}) would become fractional shear
errors of
\begin{equation}
\frac{\Delta\gamma}{\gamma} \approx \frac{2 e_\text{RMS} \Delta
  e_\text{RMS} }{1-e_\text{RMS}^2} \approx 0.01.
\end{equation}
In the context of upcoming lensing surveys that seek to constrain
shears to the per cent level or better, a systematic error of this
magnitude in shear calibration is quite serious.

Regarding possible biases in the morphological mixtures of galaxies
due to overdensities or underdensities in the training sample, there
are results in the literature for several methods that show how shear
biases vary with morphology.  For example, for the maximum likelihood
fitting code {\sc im3shape}, figure 2 in \cite{2012MNRAS.427.2711K}
shows multiplicative biases for two-component \sersic profile galaxies
as a function of their bulge-to-total ratios (denoted there as
$F_b/(F_b+F_d)$, which we will equate with our $B/T$).  The shear
calibration bias scales roughly like $0.04 - 0.05(B/T)$ as $B/T$ goes
from $0$ to $1$.  Our results suggest that typical (median) $B/T$
values may be influenced by cosmic variance in the COSMOS field
leading to fluctuations of order $0.05$.  The resulting variation in
the calibration bias would therefore be $\sim 2.5\times 10^{-3}$, or
$0.25$ per cent shear calibration uncertainty.  For existing datasets
this is not very problematic, but for surveys like LSST, Euclid, and
WFIRST-AFTA, this would be a dominant part of the systematic error
budget.  As another example, for re-Gaussianization, figure 9 of
\cite{2012MNRAS.420.1518M} shows that as \sersic $n$ goes from $1$ to
$6$, the shear calibration bias varies by $2$ per cent.  In this case,
since we have shown that the median value of \sersic $n$ can vary by
$\sim 0.4$ due to morphology-density correlations, this suggests that
the shear calibration for re-Gaussianization could be misestimated by
$\sim 2\times 10^{-3}$, or 0.2 per cent.  This too is acceptable in
existing datasets, but not those that will be used for shear
estimation in the next decade.
